---
title: "Class 09 Mini Project"
author: "Jesus (A17597539)"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"
candy <- read.csv("candy-data.csv", row.names=1)
head(candy)
candy
```

```{r}
dim(candy)
sum(candy$fruity)
```
>Q1. How many different candy types are in this dataset?

There are 85 different types of candy

>Q2. How many fruity candy types are in the dataset? 

There are 38 fruity types in this dataset


# What is your favorite candy?
```{r}
candy["Twix", ]$winpercent
candy["Starburst", ]$winpercent
candy["Kit Kat", ]$winpercent
candy["Tootsie Roll Snack Bars", ]$winpercent
```
>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy in the dataset is Starburst and its winpercent value is 67.03763

>Q4. What is the winpercent value for “Kit Kat”?

Kit Kat's winpercent value is 76.7686

>Q5.What is the winpercent value for “Tootsie Roll Snack Bars”?

Tootsie Roll Snack Bars' winpercent value is 49.6535

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)

```


>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent variable is on a different scale. The p100 of all variable are near 1 but winpercent is at 84.18

>Q7. What do you think a zero and one represent for the candy$chocolate column?

```{r}
skim(candy$chocolate)
```

A zero represents that there is no chocolate

A one represents that there is chocolate

>Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)

ggplot(candy) + aes(winpercent) +geom_histogram(binwidth=10)
```


>Q9. Is the distribution of winpercent values symmetrical?

The distribution of the winpercent values are not symmetrical

>Q10. Is the center of the distribution above or below 50%?

The center of the distribution of the histogram is below 50%


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.inds <-  as.logical(candy$chocolate)
choc.win <-  candy[choc.inds, ]$winpercent

choc.win
mean(choc.win)
```

```{r}
fru.inds <-  as.logical(candy$fruity)
fru.win <-  candy[fru.inds,]$winpercent

fru.win
mean(fru.win)
```

On average chocolate candy is ranked higher than fruity candy

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fru.win)
```
 The p-value is 2.871 e-08
 
 The difference is statistically significant because our p-value is less than 0.05
 
 #Overall Rankings
 
>Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)

candy %>% arrange(winpercent) %>% head(5)
```
The five least liked candy types are nik L nip, boston bakes beans, chiclets, super bubble, and jawbusters

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(winpercent) %>% tail(5)

```
The top 5 most liked candy types are snickers, kit kat, twix, reese's minatures, and reese's peanut butter cup

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy)+
aes(winpercent,reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets

>Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst

# Taking a look ar pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The Reese's minatures chocolate candy type is the highest ranked in terms of winpercent 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The five most expensive candy are Nik L Nip, Ring Pop, Nestle Smarties, Hershey's Krackel, and Hershey's Milk Chocolate.

The least popular of the five is Nik L Nip


>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Bar plot with pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_col()
```

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

#Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity are anit-correlated

>Q23. Similarly, what two variables are most positively correlated?

Chocolate and Winpercent are positively correlated

# Principal Component Analysis

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#install.packages("plotly")
#library(plotly)
```

```{r}
#ggplotly(p)
```
```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The original variables that are picked up strongly by PC1 are fruity, hard, and pluribus. These do make sense to me because fruity candy is typically hard and come in multiples like skittles.